---
title: "SOE data structure"
author: "Sean Lucey"
date: "October 12, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
data.dir <- '/home/slucey/EcoAP/SOE2017/data'
load(file.path(data.dir, 'SOE_data.RData'))
library(data.table)
```

The NEFSC State of the Ecosystem (SOE) report is a collection of indicators that
track the status and trend of the Northeast US Large Marine Ecosystem.  The report
is broken into two separate regional reports to be delivered to the two Fishery 
Management Councils that the Center serves: Mid-Atlantic and New England. The 
SOE contains a diverse range of indicators from revenue and commercial fisheries
reliance through biomass of aggregate species groups to primary production and
climate drivers.

## Data Format

There are two different formats for structuring data: wide format and long format.  
Most people are familiar with the wide format that structures data like this:

```{r wide data}
wide.data <- data.table(Year = c('1', '2', '3'),
                        VarA = c('1', '2', '3'),
                        VarB = c('4', '5', '6'),
                        VarC = c('7', '8', '9'))
knitr::kable(wide.data, align = 'c', caption = 'The wide format for data frames')
```

While this is convenient for reading and human interpretation, many analyses and 
graphing routines prefer the long format:

```{r long data}
long.data <- data.table(Year = rep(c('1', '2', '3'), 3),
                        Var  = c(rep('A', 3), rep('B', 3), rep('C', 3)),
                        Value = 1:9)
knitr::kable(long.data, align = 'c', caption = 'The long format for data frames')
```

The long format makes it easier to generically subset the data to be used in 
functions.  For this reason, EDAB would prefer to receive data in the long format.
More specifically, we are looking for the following columns: 

\newpage

```{r required columns}
column.table <- data.table(column = c('Time', 'Var', 'Value', 'Units', 'Region', 'Source'),
                           description = c(
                             'Time stamp of the variable.  Typically this is the year but could be day of the week or month, etc.',
                             'Name of variable.  Use a short description such as "Piscivore Landings".',
                             'The numerical value for the data point.',
                             'Units for the variable.  See standardized list below.',
                             'Ecological production unit or shelf-wide. See standardized list below.',
                             'Where did the data come from.  This will allow for reproducibility of future reports.'
                           ))
knitr::kable(column.table, align = 'l', caption = 'Required columns for the SOE')
```

### Standarized lists

Here is the list of standardized unit notation.  Note that this list may not be
comprehensive.  However this should give you an idea of how to label units.  
If the units of your variable are not list please 
contact Sean Lucey for guidance.  Having standard notation for units 
helps in labeling the y-axis.

```{r unit codes}
units <- data.table(Abbreviations = c('unitless', 'anomaly', 'degrees latitude',
                                      'km^2', 'degrees C', 'n', 'US dollars',
                                      'n per 100m3', 'kg tow^-1'),
                    Units =c('no units', 'standardized anomaly', 'position in latitude',
                             'square kilometers', 'temperature in Celsius', 
                             'number', 'currency', 'number per 100 cubic meters',
                             'kilograms per tow'))
knitr::kable(units, align = 'l', caption = 'Examples of unit standardization')
```

Please use the abbreviations below to indicate to which ecological production unit 
(EPU) your indicator belongs.  This will aid in separating the data between the
Mid-Atlantic and New England.

```{r region codes}
regions <- data.table(Abbreviation = c('GB', 'GOM', 'MAB', 'SS', 'All'),
                      EPU = c('Georges Bank', 'Gulf of Maine', 'Mid-Atlantic Bight',
                              'Scotian Shelf', 'Shelf-wide'))
knitr::kable(regions, align = 'l', caption = 'Standard notation for EPUs')
```

\newpage

### Covert from wide to long

If your data is already in wide format, here is a simply function for converting 
your data to long.  Run this code in R and follow the example.  Note there are 
some R packages that can also do this.

&nbsp;

```{r wide to long, echo = T}
#This function requires the data.table package
library(data.table)

#Function to convert from wide to long
w2l <- function(x, by, by.name = 'Time', value.name = 'Value'){
  x.new <- copy(x)
  var.names <- names(x)[which(names(x) != by)]
  out <- c()
  setnames(x.new, by, 'by')
  for(i in 1:length(var.names)){
    setnames(x.new, var.names[i], 'V1')
    single.var <- x.new[, list(by, V1)]
    single.var[, Var := var.names[i]]
    out <- rbindlist(list(out, single.var))
    setnames(x.new, 'V1', var.names[i])
  }
  setnames(x.new, 'by', by)
  setnames(out, c('by', 'V1'), c(by.name, value.name))
}

#Simple example from above
#Wide table
wide.format <- data.table(Year = c('1', '2', '3'),
                        VarA = c('1', '2', '3'),
                        VarB = c('4', '5', '6'),
                        VarC = c('7', '8', '9'))
print(wide.format)
long.format <- w2l(wide.format, by = 'Year')
print(long.format)

#Notice that the w2l function named the columns to the ones noted above
#but did not add units, region, or source
#Add units and region manually
units  <- data.table(Units  = c(rep('kg tows^-1', 6),
                                rep('n per 100m3', 3)))
region <- data.table(Region = rep('GB', 9))
source <- data.table(Source = c(rep('NEFSC survey data (Survdat)', 6),
                                rep('NEFSC ECOMON data', 3)))
long.format <- cbind(long.format, units)
long.format <- cbind(long.format, region)
long.format <- cbind(long.format, source)
print(long.format)
```

If you have any issues converting your data please contact Sean Lucey.

