---
title: "Using the Mid-Atlantic EAFM Risk Assessment: Possible Next Steps"
author:
- name: Sarah K. Gaichas, Geret S. DePiper 
  affiliation: NOAA NMFS Northeast Fisheries Science Center, Woods Hole, MA, USA
- name: Brandon W. Muffley
  affiliation: Mid-Atlantic Fishery Management Council, Dover, DE, USA
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  pdf_document:
    includes:
      in_header: riskassess-preamble-latex.tex
    keep_tex: yes
    toc: no
    template: svm-latex-ms.tex
  html_document:
    toc: yes
  word_document:
    toc: no
fontsize: 11pt
geometry: margin=0.8in
csl: frontiers.csl
bibliography: My_Library.bib
---


```{r Directory and Data Set-up, echo = F, message = F , warning = F}
#data.dir  <- '/home/slucey/EcoAP/SOE2017/data'
#image.dir <- '/home/slucey/EcoAP/SOE2017/images'
#gis.dir   <- '/home/slucey/EcoAP/SOE2017/GIS'
#data.dir <- 'Z:\\SOE2017\\data'
#image.dir <- 'Z:\\SOE2017\\images'
data.dir  <- '/users/sgaichas/Documents/0_Data/ESR/SOE2017/data'
image.dir <- '/users/sgaichas/Documents/0_Data/ESR/SOE2017/images'
gis.dir <- '/users/sgaichas/Documents/0_Data/ESR/SOE2017/GIS'
#data.dir  <- './data'
#image.dir <- './images'
#gis.dir <- './usa_shape_files'
#data.dir  <- '/home/sgaichas/Data/Projects/ESR/SOE2017/data'
#image.dir <- '/home/sgaichas/Data/Projects/ESR/SOE2017/images'
#gis.dir <- '/home/sgaichas/Data/Projects/ESR/SOE2017/GIS'
PKG <- c("data.table","Kendall","rgdal","tidyverse","nlme", 'magick',"knitr",'kableExtra','zyp')
for (p in PKG) {
  if(!require(p,character.only = TRUE)) {  
    install.packages(p)
    require(p,character.only = TRUE)}
}
load(file.path(data.dir, 'SOE_data.RData'))
load(file.path(data.dir, 'Comm_Climate_Vul.RData'))
load(file.path(data.dir, 'Fisheries_Eng_Rel.RData'))
knitr::opts_chunk$set(echo = FALSE, fig.align='center',fig.path = './riskassessfigs/')

#to get table captions and numbering
tn = local({
  i = 0
  function(x) {
    i <<- i + 1
    #paste('\n\n:Table ', i, ': ', x, sep = '')
    paste('\n\n:', x, sep = '')
    # The : before Table tells pandoc to wrap your caption in <caption></caption>
  }
})
knit_hooks$set(tab.cap = function(before, options, envir) {
  if(!before)
    tn(options$tab.cap)
})
default_output_hook = knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
  if (is.null(options$tab.cap) == F)  
    x
  else
    default_output_hook(x,options)
})

```

```{r Generic Plot Function, echo = F}
#Plot figure new
soe.plot <- function(data, x.var, y.var, x.label = '', y.label = '', tol = 0.1,
                     x.start = NA, x.end = NA, end.start = 2006, bg.col = background,
                     end.col = recent, trend.neg = main.neg, trend.pos = main.pos,
                     end.trend.neg = end.neg, end.trend.pos = end.pos,

                     stacked = NA, x.line = 2.6, y.line = 3.5, scale.axis = 1,
                     rel.y.num = 1.5, rel.y.text = 1.5){
 
  #Select Data
  x <- data[Var == y.var, ]
  x <- x[order(x[, get(x.var)]), ]
  setnames(x, x.var, 'X')
 
  #Set common time step if necessary
  if(is.na(x.start)) x.start <- min(x[, X])
  if(is.na(x.end))   x.end   <- max(x[, X])
  x <- x[X >= x.start, ]
 
  #Set up plot parameters
  y.max <- max(x[, Value], na.rm = T) + tol * max(x[, Value], na.rm = T)
  y.min <- min(x[, Value], na.rm = T) - tol * abs(min(x[, Value], na.rm = T))
  y.mean <- mean(x[, Value], na.rm = T)
  y.sd <- sd(x[, Value], na.rm = T)
 
  #Plot blank plot
  plot(x[X >= x.start, list(X, Var)], xlim = c(x.start, x.end),
       ylim = c(y.min,y.max), xlab = '', ylab = '', axes = F, ty = 'n')
 
  #Add background
  u <- par('usr')
  rect(u[1], u[3], u[2], u[4], border = NA, col = bg.col)
 
  #Add end period shading
  rect(end.start - 0.5, u[3], u[2], u[4], border = NA, col = end.col)
  abline(h = u[3], lwd=3)
  abline(v = u[1], lwd=3)
 
  #Add grides
  abline(h = y.mean + y.sd, col = 'white', lwd = 3, lty = 2)
  abline(h = y.mean - y.sd, col = 'white', lwd = 3, lty = 2)
 
  #Add data points/lines
  points(x[, list(X, Value)], pch = 16, cex = 1.5)
  lines( x[, list(X, Value)], lwd = 2)
 
  #Add axis
  if(is.na(stacked)) axis(1, cex.axis = 1.5)
  if(!is.na(stacked)){
    if(stacked!= 'A') axis(3, cex.axis = 1.5, tck = 0.1, labels = F)
  }
  #Stacked axes with 0 overlap so need to remove
  if(scale.axis != 1){
        labels <- axTicks(2) / scale.axis
        if(labels[1] == 0) labels[1] <- ''
        axis(2, at = axTicks(2), labels = as.numeric(labels), cex.axis = rel.y.num,
             las = T)
    } else {axis(2, cex.axis = rel.y.num, las = T)}

  #Identify significant trends
  #Whole time series
  mksts <- zyp.trend.vector(x[, Value])
  mkstsp <- round(unlist(mksts[6]), 3)
 
  if(mkstsp < 0.05){
    lmod <- lm(x[, Value] ~ x[, X])
    lmod_c <- unlist(lmod[1])
    lmod_i <- lmod_c[1]
    lmod_s <- lmod_c[2]
    if(lmod_s > 0){
      lines(x[, X], lmod_s * x[, X] + lmod_i,
            col = trend.pos, lty = 1, lwd = 7)
    }
    if(lmod_s < 0){
      lines(x[, X], lmod_s * x[, X] + lmod_i,
            col = trend.neg, lty = 1, lwd = 7)
    }
  }
 
  #Final portion of time series
  mksld <- zyp.trend.vector(x[X > (end.start - 1), Value])
  mksldp <- round(unlist(mksld[6]), 3)
  if(mksldp < 0.05){
    l10_x <- x[X > (end.start - 1), X]
    l10_y <- x[X > (end.start - 1), Value]
    lmod <- lm(l10_y ~ l10_x)
    lmod_c <- unlist(lmod[1])
    lmod_i <- lmod_c[1]
    lmod_s <- lmod_c[2]
    if(lmod_s > 0){
      lines(l10_x, lmod_s * l10_x + lmod_i,
            col = end.trend.pos, lwd=7)
    }
    if(lmod_s < 0){
      lines(l10_x, lmod_s * l10_x + lmod_i,
            col = end.trend.neg, lwd=7)
    }
  }
 
  #Add axis labels
  if(!is.na(stacked)) text(u[1], u[4], labels = stacked, cex = 2, adj = c(-0.5, 1.5))
  if(is.na(stacked)){
    mtext(1, text = x.label, line = x.line, cex = 1.5)
    mtext(2, text = y.label, line = y.line, cex = rel.y.text)
  }
}

#Add axis labels for stacked plots
soe.stacked.axis <- function(x.label, y.label, x.line = 2.6,
                             y.line = 3.5, rel.y.text = 1.5){
  axis(1, cex.axis = 1.5)
  mtext(1, text = x.label, line = x.line, cex = 1.5, outer = T)
  mtext(2, text = y.label, line = y.line, cex = rel.y.text, outer = T)
}

```


```{r Plot options, echo = F}
#Background colors
background   <- '#F4F7F2'
recent       <- '#E6E6E6'
#trend lines
main.pos <- rgb(253/255, 184/255, 99/255,  alpha = 0.8)
main.neg <- rgb(178/255, 171/255, 210/255, alpha = 0.8)
end.pos  <- rgb(230/255, 97/255,  1/255,   alpha = 0.8)
end.neg  <- rgb(94/255,  60/255,  153/255, alpha = 0.8)

```

#Introduction
The Council approved an EAFM Guidance Document in 2016 which outlined a path forward to more fully incorporate ecosystem considerations into marine fisheries management (http://www.mafmc.org/s/EAFM_Guidance-Doc_2017-02-07.pdf). The Council’s stated goal for EAFM is "to manage for ecologically sustainable utilization of living marine resources while maintaining ecosystem productivity, structure, and function." Ecologically sustainable utilization is further defined as "utilization that accommodates the needs of present and future generations, while maintaining the integrity, health, and diversity of the marine ecosystem." Of particular interest to the Council was the development of tools to incorporate the effects of species, fleet, habitat and climate interactions into its management and science programs. To accomplish this, the Council agreed to adopt a structured framework to first prioritize ecosystem interactions, second to specify key questions regarding high priority interactions and third tailor appropriate analyses to address them [@gaichas_framework_2016]. Because there are so many possible ecosystem interactions to consider, a risk assessment was adopted as the first step to identify a subset of high priority interactions [@holsman_ecosystem-based_2017]. The risk elements included in the Council's initial assessment spanned biological, ecological, social and economic issues (Table \ref{riskel}) and risk criteria for the assessment were based on a range of indicators and expert knowledge (Table \ref{allcriteria}).

This document outlines potential next steps to follow up on the Mid-Atlantic Council’s initial EAFM risk assessment. The risk assessment was designed help the Council decide where to focus limited resources to address ecosystem considerations by first clarifying priorities. Overall, the purpose of the EAFM risk assessment is to provide the Council with a proactive strategic planning tool for the sustainable management of marine resources under its jurisdiction, while taking interactions within the ecosystem into account. 

#Discussion
<!--Because this was the initial EAFM risk assessment for the Council in what is intended to be a continuing process rather than a static product, we were able to identify areas to improve in future years as the process is repeated. For example, the short timeframe for completing this risk assessment required that we postpone work on statistical methods for indicator selection [@otto_quantitative_2017; @van_oudenhoven_framework_2012]. However, many elements seleted by the Council had only one indicator time series available, such that selection methods were not required. Nevertheless, further testing of the indicators for consistency and reliability [@samhouri_quantitative_2009; @fay_testing_2013; @shin_specificity_2018] should be a high priority for use in future assessments. In addition, statistical analysis of time series requires further examination. We used pre-whitening to address potential autocorrelation in indicator time series that could result in unreliable statistical significance tests [@yue_influence_2002]. However, further simulation work is required to determine the reliability of significance tests for time series indicators with different degrees of autocorrelation and trend strength (see, e.g., @planque_principal_2018). Given the flexible nature of the risk assessment, any necessary updates can be incorporated in future iterations. Finally, we reiterate that for many of these elements (aside from biomass and fishing mortality status), there are no established risk thresholds, so the consensus of the Council and committee were used to establish risk levels. If clear thresholds become available for any element/indicator combination, in particular through further indicator analysis [@large_defining_2013; @fay_management_2015] they can be substituted into future assessments. -->

Risk assessment provides a starting point for prioritizing further, more detailed analysis. An indicator-based risk assessment does not provide a mechanistic assessment of the system, which is only possible through a structural modeling approach, which could be focused specifically on fully assessing and mitigating identified risks. According to the Council's accepted framework within the EAFM policy guidance, risk assessment results will be used to prioritize and refine the key management questions and issues for further evaluation [@gaichas_framework_2016]. The next step will be the development of example conceptual models which aim to showcase different approaches the Council could use to move from this initial risk assessment to specifying a more in-depth analysis of particular high prority questions. 

One potential prioritization approach focuses on the single species with the largest number of high risk rankings across all elements, which is black sea bass (8 high Risk Elements, Tables \ref{sptable}-\ref{ecotable}). A second approach focuses on the fishery with the highest landings value (as a proxy for seafood production and economic benefits), which is the squid complex, with uncertainties in assessment and potential for distribution shifts along with high risks identified across multiple management elements (Tables \ref{sptable}-\ref{spsectable}). A third approach focuses on the risk element with the most high risk categorizations across all categories, which is allocation (12 high risk species/sector combinations; Table \ref{spsectable}). These prioritization approaches use only the risk elements that could be ranked; inclusion of additional risk elements of interest to the Council (Table \ref{riskelpa}) may not change the selected approach but could change the outcome.

An example conceptual model based on the first prioritization approach has been developed to illustrate the process (Fig. \ref{bsbex}).  This model was developed by modifying the existing Mid-Atlantic conceptual model using the risk assessment results tables to ensure that all key factors affecting black sea bass and its fisheries were considered. We can step through this process briefly during the EOP presentation to illustrate methods for building a model. While it looks complex, this need not be time consuming and the process can incorporate the knowledge of managers, fishery participants, and scientsts to ensure that key interactions and management priorities are considered and accounted for. 

A (partial) conceptual model based on the second approach was recently developed as well, which emphasizes habitat linkages affecting availability for longfin squid as well as economic and market factors affecting fishery effort (Fig. \ref{squidex}). Again, this is illustrative of an approach and not intended to be prescriptive. 

In the Council EAFM framework, specifying a conceptual model in turn provides the initial scoping for more in-depth management strategy evaluation (MSE), where simulation modeling is used to test alternative methods for achieving specific management objectives under uncertainty [@sainsbury_design_2000; @butterworth_why_2007; @punt_developing_2007; @punt_management_2016]. Any MSE undertaken by the Council will be a deliberative process and will include a variety of stakeholder and Council engagement/involvement. MSE is generally recognized as a core component of the ecosystem approach [@smith_scientific_2007; @levin_integrated_2009; @levin_guidance_2014], but because it necesessitates a high level of engagement, MSE can be a resource-intensive process when applied even to relatively simple single species harvest control rules [e.g., @jones_stakeholder-centered_2016]. EAFM potentially expands the scope of interactions and uncertainties to consider in evaluating harvest control rules and other fishery management procedures. How can managers ensure that the most important MSE gets done first? Risk assessment provides a systematic framework to ensure that limited MSE resources address the highest priority ecosystem interactions and risks. A risk matrix, such as the one developed by the Council, can be used to quickly evaluate where further integrated analysis and MSE should be focused--which fishery management plans, which species, and which Risk Elements need to be included in the analysis. Further, because the risk assessment includes social and economic elements as well as the more standard ecological elements, the conceptual model and resulting MSE are designed to include the relevant linkages and therefore avoid unintended consequences of management actions [@degnbol_unintended_2007]. 

<!--Similar to many other regions worldwide, natural resource management challenges are mounting in the Mid-Atlantic, from climate change to increasing needs for food and recreational opportunities to increasing uses of ocean areas (e.g. wind farm development). The process described here was completed within the current management system without displacing other processes, and adds value now and for future planning. This EAFM risk assessment highlights certain species and certain management issues as posing higher cumulative risks to meeting Council-derived management objectives when considering a broad range of ecological, social, and economic factors. The risk assessment result summaries will be used by the Council to prioritize further EAFM analyses as well as develop research plans and Council planning priorities over the coming five years. As ecosystem reporting and operational EAFM continue to evolve in future years, the Council foresees refining the process so that ecosystem indicators monitor risks to achieving ecological, social, and economic fishery objectives, which can then be mitigated through management action. Overall, ecosystem indicator-based risk assessment is a method that can be adapted to a wide range of resource management systems and available information, and therefore a promising way forward in the implementation of EAFM.-->

<!--
#Data Availability Statement
Data sources are listed within the manuscript as references if published or with URL links if online. State of the Ecosystem indicator data used by the Northeast Fisheries Science Center will be available at the IEA website via an ERDDAP server startin in August 2018.
-->

#References
<div id="refs"></div>

\newpage
#Figures

```{r, fig.cap="Example conceptual model for a species determined to be high risk: black sea bass. Note that this is an example for discussion and not a Council analysis. \\label{bsbex}", fig.align='center', out.width="\\linewidth", echo=F, message=F}

image_read(file.path(image.dir,"BSBcentric2.png")) %>%
image_trim()

```

\newpage
\blandscape

```{r, fig.cap="Example conceptual model for a fishery determined to be high risk: longfin squid. Note that this is an example for discussion and not a Council analysis. \\label{squidex}", fig.align='center', out.width="\\linewidth" , echo=F, message=F}

image_read(file.path(image.dir,"wintersquid2qpress.png")) %>%
image_trim()

```
\elandscape

\newpage


#Tables
```{r riskel, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}

#tab.cap="Risk Elements, Definitions, and Indicators Used\\label{riskel}",

elem <-read.table("riskelements.txt", sep="|", header=F, strip.white = T, stringsAsFactors = F)
elem <- elem[,2:4]
names(elem) <- c("Element", "Definition", "Indicator")

# elem$Element <- factor(all$Element, levels=c("Assessment performance", "F status", "B status", "Food web  (Council Predator)", "Food web  (Council Prey)", "Food web  (Protected Species Prey)",
#                                             "Ecosystem productivity", "Climate", "Distribution shifts", "Estuarine habitat", "Offshore habitat", "Commercial Revenue",
#                                             "Recreational Angler Days/Trips", "Commercial Fishery Resilience (Revenue Diversity)", "Commercial Fishery Resilience (Shoreside Support)",
#                                             "Fleet Resilience", "Social-Cultural", "Commercial", "Recreational", "Control", "Interactions", "Other ocean uses", "Regulatory complexity", 
#                                             "Discards", "Allocation"))

kable(elem, format = "latex", booktabs = T, longtable=T, caption="Risk Elements, Definitions, and Indicators Used\\label{riskel}") %>%
  kable_styling(font_size=8, latex_options=c("repeat_header")) %>%
  column_spec(1, width="2.5cm") %>%
  column_spec(2:3, width="7cm") %>%
  group_rows("Ecological",1,11) %>%
  group_rows("Economic",12,15) %>%
  group_rows("Social",16,17) %>%
  group_rows("Food Production",18,19) %>%
  group_rows("Management",20,25)
  #landscape()


```
 

\newpage  
 
```{r allcriteria,  echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}

#tab.cap="Risk Ranking Criteria used for each Risk Element\\label{allcriteria}",

all<-read.table("riskrankingcriteria.txt", sep="|", header=T, strip.white = T, stringsAsFactors = F)
names(all) <- c("Element", "Ranking", "Criteria")
all$Ranking <- factor(all$Ranking, levels=c("Low", "Low-Moderate",  "Moderate-High", "High"))
all$Element <- factor(all$Element, levels=c("Assessment performance", "F status", "B status", "Food web  (MAFMC Predator)", "Food web  (MAFMC Prey)", "Food web  (Protected Species Prey)",
                                            "Ecosystem productivity", "Climate", "Distribution shifts", "Estuarine habitat", "Offshore habitat", "Commercial Revenue",
                                            "Recreational Angler Days/Trips", "Commercial Fishery Resilience (Revenue Diversity)", "Commercial Fishery Resilience (Shoreside Support)",
                                            "Fleet Resilience", "Social-Cultural", "Commercial", "Recreational", "Control", "Interactions", "Other ocean uses", "Regulatory complexity", 
                                            "Discards", "Allocation"))
allwide <- all %>% 
  spread(Ranking, Criteria)
kable(allwide, format = "latex", booktabs = T, longtable=T, caption="Risk Ranking Criteria used for each Risk Element\\label{allcriteria}") %>%
  kable_styling(font_size=8, latex_options=c("repeat_header")) %>%
  column_spec(1, width="2cm") %>%
  column_spec(2:5, width="5cm") %>%
  landscape()


```
 

\newpage

```{r sptable,  echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}

#tab.cap="Species level risk analysis results; l=low risk (green), lm= low-moderate risk (yellow), mh=moderate to high risk (orange), h=high risk (red)\\label{sptable}",

# spplist     oc,  sc,  flk, scp, bsb, mack, but, lsq, ssq, gtile,  btile,  blu, dog, monk
risk.species<-data.frame(
  Species = c("Ocean Quahog", "Surfclam", "Summer flounder", "Scup", "Black sea bass", "Atl. mackerel", "Butterfish", "Longfin squid", "Shortfin squid", "Golden tilefish", "Blueline tilefish", "Bluefish", "Spiny dogfish", "Monkfish", "Unmanaged forage", "Deepsea corals"),
  Assess  = c("l", "l", "l",  "l", "l", "l",  "l", "lm", "lm", "l", "h", "l", "lm", "h", "na", "na"),
  Fstatus = c("l", "l", "h", "l", "l", "h", "l", "lm", "lm", "l", "h", "l", "l", "lm", "na", "na"),
  Bstatus = c("l", "l", "lm", "l", "l", "h", "l", "lm", "lm", "lm", "mh", "lm", "lm", "lm", "na", "na"),
  FW1Pred = c("l", "l", "l", "l", "l", "l", "l", "l", "l", "l", "l", "l", "l", "l", "l", "l"),
  FW1Prey = c("l", "l", "l", "l", "l", "l", "l", "l", "l", "l", "l", "l", "l", "l", "lm", "l"),
  FW2Prey = c("l", "l", "l", "l", "l", "l", "l", "lm", "lm", "l", "l", "l", "l", "l", "lm", "l"),
  Climate = c("h", "mh", "lm", "lm", "mh", "lm", "l", "l", "l", "mh", "mh","l", "l", "l", "na", "na"),
  DistShift = c("mh", "mh", "mh", "mh", "mh", "mh", "h", "mh", "h", "l", "l", "mh", "h", "mh", "na", "na"),
  EstHabitat = c("l", "l", "h", "h", "h", "l", "l", "l", "l", "l", "l", "h", "l", "l", "na", "na")#,
#  OffHabitat = c("na", "na", "l", "l", "l", "l", "l", "l", "h", "na", "na", "na", "l", "l", "na", "na")#,
)

# these elements were removed by the council
#  PopDiv = c("na", "na", "na", "na", "na", "na", "na", "na", "na", "na", "na", "na", "na", "na"),
#  FoodSafe = c(),

# one column test
# risk.species %>%
#   mutate(Fstatus = 
#     cell_spec(Fstatus, format="latex", color = "black", align = "c", background =factor(Fstatus, c("na", "l", "lm", "mh", "h"),c("white", "green", "yellow", "orange", "red")))) %>%
#   kable(risk.species, format="latex", escape = F, booktabs = T, linesep = "")

#generalize to all
risk.species %>%
  mutate_at(vars(-Species), function(x){ 
    cell_spec(x, format="latex", color = "gray", align = "c", background =factor(x, c("na", "l", "lm", "mh", "h"),c("white", "green", "yellow", "orange", "red")))}) %>%
  kable(risk.species, format="latex", escape = F, booktabs = T, linesep = "", 
        caption="Species level risk analysis results; l=low risk (green), lm= low-moderate risk (yellow), mh=moderate to high risk (orange), h=high risk (red)\\label{sptable}") %>%
  kable_styling(latex_options = "scale_down")


```

\newpage

```{r spsectable,  echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
#tab.cap="Species and sector level risk analysis results; l=low risk (green), lm= low-moderate risk (yellow), mh=moderate to high risk (orange), h=high risk (red)\\label{sptable}",

risk.sppsector<-data.frame(
  Species = c("Ocean Quahog-C", "Surfclam-C", "Summer flounder-R", "Summer flounder-C","Scup-R", "Scup-C","Black sea bass-R", "Black sea bass-C","Atl. mackerel-R", "Atl. mackerel-C","Butterfish-C", "Longfin squid-C", "Shortfin squid-C", "Golden tilefish-R", "Golden tilefish-C","Blueline tilefish-R","Blueline tilefish-C", "Bluefish-R", "Bluefish-C","Spiny dogfish-R", "Spiny dogfish-C", "Unmanaged forage", "Deepsea corals"),
  MgtControl = c(1,1,3,2,1,1,4,2,1,1,1,1,1,9,1,1,1,2,1,1,1,9,9),
  TecInteract = c(1,1,1,3,1,3,1,2,1,2,2,3,2,1,1,1,1,1,1,1,3,9,9),
  OceanUse = c(2,2,2,2,2,2,3,4,1,3,3,4,2,1,1,1,1,1,2,1,3,9,3),
  RegComplex = c(1,1,4,3,3,3,4,3,1,4,4,4,2,1,1,3,3,1,2,1,3,9,9),
  Discards = c(1,1,4,2,3,3,3,2,1,2,3,4,1,1,1,1,1,3,2,1,2,9,9),
  Allocation = c(1,1,4,4,1,1,4,4,4,4,1,4,1,1,1,4,4,4,4,1,4,9,9) 
)

#convert to text for consistency
risk.sppsector <- risk.sppsector %>%
     mutate_at(vars(-Species), function(x){
       recode(x,'1'="l",'2'="lm",'3'="mh",'4'="h",'9'="na")}) %>%
     as.data.frame()

#make table
risk.sppsector %>%
  mutate_at(vars(-Species), function(x){ 
    cell_spec(x, format="latex", color = "gray", align = "c", background =factor(x, c("na", "l", "lm", "mh", "h"),c("white", "green", "yellow", "orange", "red")))}) %>%
  kable(risk.sppsector, format="latex", escape = F, booktabs = T, linesep = "",
        caption="Species and sector level risk analysis results; l=low risk (green), lm= low-moderate risk (yellow), mh=moderate to high risk (orange), h=high risk (red)\\label{spsectable}") %>%
  kable_styling(font_size =  9)

```

\newpage

```{r ecotable,  echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
  
#tab.cap="Ecosystem level risk analysis results; l=low risk (green), lm= low-moderate risk (yellow), mh=moderate to high risk (orange), h=high risk (red)\\label{sptable}",

risk.eco<-data.frame(
  System = c("Mid-Atlantic"),
  EcoProd = c("lm"),
  #EcoDiv = c("lm"),
  CommRev = c("mh"),
  RecVal = c("h"),
  FishRes1 = c("l"),
  FishRes4 = c("mh"),
  #CommJobs = c("mh"),
  #RecJobs = c("l"),
  FleetDiv = c("mh"),
  Social = c("lm"),
  ComFood = c("h"),
  RecFood = c("mh")
)

#make table
risk.eco %>%
  mutate_at(vars(-System), function(x){ 
    cell_spec(x, format="latex", color = "gray", align = "c", background =factor(x, c("na", "l", "lm", "mh", "h"),c("white", "green", "yellow", "orange", "red")))}) %>%
  kable(risk.eco, format="latex", escape = F, booktabs = T, linesep = "",
        caption="Ecosystem level risk analysis results; l=low risk (green), lm= low-moderate risk (yellow), mh=moderate to high risk (orange), h=high risk (red)\\label{ecotable}") %>%
  kable_styling(latex_options = "scale_down")

```

<!--
\newpage 

```{r suppriskel, tab.cap="Risk Elements, Definitions, and Indicators put aside for future evaluation\\label{riskelpa}", echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl <- "
| Risk Element         | Definition: Risk to what?                        |  Indicators used                                 |
|:-------------------------------------|:-------------------------------------------------|:-------------------------------------------------|
| **Put Aside** |  |
| Population diversity  | Risk of not achieving OY due to reduced diversity   | Size composition, sex ratio, genetic diversity |
| Ecological diveristy  | Risk of not achieving OY due to reduced diversity   | Fishery independent species diversity |
| Fishery Resilience (2)  | Risk of reduced fishery business resilience due to access to capital |  No current indicator avilable |         
| Fishery Resilience (3)  | Risk of reduced fishery business resilience due to insurance availabilty  |  No current indicator available |
| Fishery Resilience (5)  | Risk of reduced fishery business resilience due to access to emerging markets/opportunities  | Needs clarification  |
| Commercial Employment   | Risk of not optimizing employment opportunities | EOP Committee unconfident in Fisheries of US employment inicator |
| Recreational Employment | Risk of not optimizing employment opportunities | EOP Committee unconfident in Fisheries of US employment indicator |
| Seafood safety   | Risk of not maintaining market access, human health    | Number of public advisories by species |

"
cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
```
